On the Functionality of the Multilayer Perceptron Classifier Neural Network Architecture

### 2020-06-03 Thesis/MLP Notes/Mutlilayer Perceptron.tex
2 June 2020
- Updates to MLP Paper

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{multicol}
\usepackage[left=2.5cm,right=2.5cm,top=2.5cm]{geometry}

@@

\section*{Architecture}
\paragraph*{}To build a Multilayer Perceptron (MLP), artificial neurons need to be grouped into multiple structures, each called a \textit{layer}. A single layer is modeled mathematically as a column vector, $\vec{x}$ by convention:

@@

Where $b^{(l)}_i$ is the bias entry, $w^{(l)}_{ij}$ is the weight coefficient that joins neuron $x_j$ to neuron $x_i$,m adn $x_j$ is the activation of the $j$-th neuron.  Notice how the combination of elements in layer $(l)$, given by the upper index, combine to produce the $i$-th neuron activation in layer $(l+1)$.

@@

\paragraph*{}To make this more convenient to read and express mathematically, the entirety of the system (\ref{single element}) can be expresses as a single matrix vector product. The elements $x^{(l+1)}_i$ and $x^{(l)}_j$ can be concatenated in column vectors, $\vec{x}^{(l+1)}$ and $\vec{x}^{(l)}$ respectively:

@@

\begin{equation}
\label{bias}
\vec{b}^{(l)} = \big[ b^{(l)}_0 , b^{(l)}_1 , b^{(l)}_2 , ... , x^{(l)}_{n-2} , b^{(l)}_{n-1} \big]^T \\
\end{equation}

@@

\begin{equation}
\label{mat-vec + act}
\vec{x}^{(l+1)} = f^{(l)} \Big( \hat{W}^{(l)} \vec{x}^{(l)} + \vec{b}^{(l)} \Big)
\end{equation}

Where $f^{(l)}$ is the activation function for the $l$-th layer. $\vec{x}^{(l+1)}$ has shape $n \times 1$ , $\hat{W}^{(l)}$ has shape $n \times m$, $\vec{x}^{(l)}$ has shape $m \times 1$ and $\vec{b}^{(l)}$ has shape $n \times 1$. (For conventional reasons, the baias vector has been moved to the other side of the matrix-vector product.)

\paragraph*{}As the upper indexes imply, this is how the activations in the next layer are produced. By repeating this equation $L-1$ times in an 
$L$-layered network, information from the feature vector $\vec{x}^{(0)}$ can be successively transformed into the final decision output of the network $\vec{x}^{(L-1)}$. Each application of equation (\ref{mat-vec + act}) produces activations in each of the hidden layers. Note that in this $L$-layer network, there are $L-1$ weight matrices and $L-1$ bias vectors as well. Every single element in all of these arrays can be concatenated to form a larger object, $\hat{\theta}$. It is the combination of the elements in $\hat{\theta}$ that allow the network to map an input vector to an output vector as expressed in equation (\ref{approx}).

@@

\paragraph*{}To do this, we must treat the behavior of a neural network as an optimization problem. The goal of this would then be to find the set of parameters $\hat{\theta}$ that produces the smallest possible output of some \textit{error function}, \cite{Goodfellow,James,Loy}. This may also be called a \textit{cost function} or \textit{loss function} as well. This function , $J$ takes the predicted output of the network $\vec{x}^{(L-1)}$ and the parameters $\hat{\theta}$ and maps it to a real number, which then measures the \textit{cost} of that particular sample such that:

\begin{equation}
\label{cost func}
J : \big\{ \vec{x}^{(L-1)}} \rightarrow \big\{ \mathbb{R} \big\}
\end{equation}

\paragraph*{}The larger the value outputted from this function $J$, the further the models prediction differed from the expected value. Thus, this function can be through of as a measurement of how \textit{poor} the network's prediction was. It then becomes appropriate to make the goal of the model to \textit{optimize} $J$, with respect the parameters $\hat{\theta}$ to allow for a generally low cost value for a given input.

\paragraph*{}There are many ways to optimize the parameters in a neural network,for this work, we will examine an optimizer algorithm called \textit{Stochastic Gradient Descent} (SGD). This is an iterative method that uses repeated a passes over a data set to slowly push the cost function to a smaller and smaller value. As the name implies, we compute the gradient of the cost function, which its the partial derivative of each element that makes up the function, at a given step.

\paragraph*{}To compute the gradient of the cost function, we have to examine it's composition.

@@

\section*{Conclusion}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[left=2.5cm,right=2.5cm,top=2.5cm]{geometry}
\usepackage{fancyhdr}

\pagestyle{fancy}
\fancyhf{}
\lhead{MLP Classifier Neural Network}
\rhead{Landon Buell}
\cfoot{\thepage}

@@

\section*{Architectural Considerations}
\paragraph*{}To build a Multilayer Perceptron (MLP), artificial neurons need to be grouped into multiple structures, each called a \textit{layer}. A single layer is modeled mathematically as a column vector, $\vec{x}$ by convention:

@@

Where $b^{(l)}_i$ is the bias entry, $w^{(l)}_{ij}$ is the weight coefficient that joins neuron $x_j$ to neuron $x_i$, and $x_j$ is the activation of the $j$-th neuron.  Notice how the combination of elements in layer $(l)$, given by the upper index, combine to produce the $i$-th neuron activation in layer $(l+1)$.

@@

\paragraph*{}To make this more convenient to read and express mathematically, the entirety of the system (\ref{system elements}) can be expressed as a single matrix vector product. The elements $x^{(l)}_j$ and $x^{(l+1)}_i$ can be concatenated in column vectors, $\vec{x}^{(l)}$ and  $\vec{x}^{(l+1)}$ respectively:

@@

\begin{equation}
\label{bias}
\vec{b}^{(l)} = \big[ b^{(l)}_0 , b^{(l)}_1 , b^{(l)}_2 , ... , b^{(l)}_{n-2} , b^{(l)}_{n-1} \big]^T \\
\end{equation}

@@

\begin{equation}
\label{mat-vec + act}
\vec{x}^{(l+1)} = f^{(l)} \Big( \hat{W}^{(l)} \vec{x}^{(l)} + \vec{b}^{(l)} \Big)
\end{equation}
For conventional reasons, the bias vector has been moved to the other side of the matrix-vector product. This is also often represented as two separate processes:

\begin{itemize}
\item[•] The \textit{Pre-activation} linear portion:
\begin{equation}
\label{pre-act}
\vec{z}^{(l)} = \hat{W}^{(l)} \vec{x}^{(l)} + \vec{b}^{(l)}
\end{equation}
\item[•] The \textit{final-activation} portion:
\begin{equation}
\label{fin-act}
\vec{x}^{(l+1)} = f^{(l)} \Big( \vec{z}^{(l)} \Big)
\end{equation}
\end{itemize}

Where $f^{(l)}$ is the activation function for the $l$-th layer. $\vec{x}^{(l+1)}$ has shape $n \times 1$ , $\hat{W}^{(l)}$ has shape $n \times m$, $\vec{x}^{(l)}$ has shape $m \times 1$ and $\vec{b}^{(l)}$ has shape $n \times 1$.  Note that the inclusion of an activation function by layer, $f^{(l)}$ in eqn. (\ref{mat-vec + act}) changes the linear combination in eqn. (\ref{single element}) to instead become:

\begin{equation}
\label{single element + act}
x^{(l+1)}_i = f^{(l)} \Big( b^{(l)}_i + \sum_{j} w^{(l)}_{ij} x^{(l)}_j \Big)
\end{equation}

\paragraph*{}As the upper indexes imply, this is how the activations in the \textit{next} sequential layer are produced. By repeating this equation $L-1$ times in an $L$-layered network, information from the feature vector $\vec{x}^{(0)}$ can be successively transformed into the final decision output of the network $\vec{x}^{(L-1)}$. Each application of equation (\ref{mat-vec + act}) produces activations in each of the hidden layers. Note that in this $L$-layer network, there are $L-1$ weight matrices and $L-1$ bias vectors as well. Every single element in all of these arrays can be concatenated to form a larger object, $\hat{\theta}$. It is the combination of the elements in $\hat{\theta}$ that allow the network to map an input vector to an output vector as expressed in equation (\ref{approx}).

\paragraph*{}For a fully connected multilayer perceptron model with $L$ layers, we can construct an algorithm to execute the forward pass process. The algorithm returns two array-like object containing all final activations (\ref{mat-vec + act}), and pre-activation function (\ref{mat-vec}) values. The final object in the list of the output of the network $\vec{x}^{(L-1)}$ (also referred to as $y^*$). For a pseudo-code execution, see alg.(\ref{forward-pass}) \cite{Goodfellow}. Note that this algorithm uses features from a single training samples, a practical implementation should include batches of samples.

\begin{algorithm}
\label{forward-pass}
\caption{Feed-Forward System in a fully-connected Multilayer perceptron model for a single sample.}
\begin{algorithmic}

\REQUIRE Network depth of $L$ layers
\REQUIRE Object of weighting matrices: $\hat{W}^{(i)}$ with $i \in [0,1,...,L-3,L-2]$
\REQUIRE Object of bias vectors: $\vec{b}^{(i)}$ with $i \in [0,1,...,L-3,L-2]$
\REQUIRE Object of activation functions: $\vec{f}^{(i)}$ with $i \in [0,1,...,L-3,L-2]$
\REQUIRE Raw input to network: $\vec{a}$

final-activations $\leftarrow \{\}$ \\
pre-activation $\leftarrow \{\}$	\\
$x^{(0)} \leftarrow \vec{a}$ 		\\

\FOR{$l = 0,1,2,...,L-2,L-1$} 
\item $\vec{z}^{(l)} \leftarrow \vec{b}^{(l)} + \hat{W}^{(l)} \vec{a}^{(l)}$
\item $\vec{x}^{(l+1)} \leftarrow f^{(l)} \big( \vec{z}^{(l)} \big)$
\item pre-activations.insert($\vec{z}^{(l)}$)
\item final-activations.insert($\vec{x}^{(l+1)}$)
\ENDFOR
\RETURN pre-activations
\RETURN final-activations
\end{algorithmic}
\end{algorithm}

\paragraph*{}Both arrays  that are returned by the alg. (\ref{forward-pass}) contain $L-1$ elements. The $j$-th element in the \textit{final-activations} array is the element produced by applying an activation function to the $j$-th element of the \textit{pre-activations} array. These correspond to layers $l+1$ and $l$ respectively.

@@

\paragraph*{}We construct the object $\hat{\theta}$ to be some multidimensional array-like object that contains all of the parameters in the model. very generally, it can be thought of as a concatenation of all weighting and bias elements for all of the layers.

\begin{equation}
\label{theta}
\hat{\theta} \equiv \big[ \hat{W}^{(0)} , \vec{b}^{(0)} , \hat{W}^{(1)} , \vec{b}^{(1)} ,
							... , \hat{W}^{(L-2)} , \vec{b}^{(L-2)} 	\big]
\end{equation}

Where $\hat{W}^{(l)}$ and $\vec{b}^{(l)}$ are given in eqns. (\ref{weights}) and (\ref{bias}) respectively. For smaller networks, $\hat{\theta}$ may contain several hundred or thousand elements, and for larger deep-learning networks, may contain tens to hundreds of thousands of elements \cite{Goodfellow,Loy}. It is the combinations of these elements that allow for some input given $\vec{x}^{(0)}$ to be transformed into an output $\vec{x}^{(L-1)}$.

\paragraph*{}To do this, we treat the behavior of a neural network as a multidimensional optimization problem. The goal of this would then be to find the set of parameters $\hat{\theta}$ that produces the smallest possible output of some \textit{error function}, \cite{Goodfellow,James,Petrik}. This may also be called a \textit{cost function} or \textit{loss function} as well. This function , $J$, takes the output of the network $\vec{x}^{(L-1)}$ (which we will refer to as $y^*$) to represent a predicted or approximated output) on a given training sample, and compares it to the expected output for that sample $\vec{y}$, and maps it to a real number, which then measures the \textit{cost} of that particular sample such that:

\begin{equation}
\label{cost func}
J : \big\{ y^* , y  \big\} \rightarrow \big\{ \mathbb{R} \big\}
\end{equation}

\paragraph*{}The larger the value outputted from this function $J$, the further the model's prediction differs from the expected value. Thus, this function can be through of as a measurement of how \textit{poor} the network's prediction was. It then becomes appropriate to make the goal of the neural network to \textit{optimize} $J$, with respect the parameters $\hat{\theta}$ to allow for a generally low cost value for a given input.

\paragraph*{}There are many ways to optimize the parameters in a neural network,for this work, we will examine an optimizer algorithm called \textit{Stochastic Gradient Descent} (SGD). This is an iterative method that uses repeated a passes over a data set to slowly push the cost function to a smaller and smaller value \cite{Geron,Goodfellow,Loy}. As the name implies, we compute the gradient of the cost function with respect to all of the parameters in the model to produce a vector. The components of this vector give the direction to which a small change the elements of $\hat{\theta}$ will result in a small reduction of the value of the loss function. Recall the gradient vector of some function $h$ of variables $x_0$ , $x_1$ , ... ,$x_{n-2}$ , $x_{n-1}$ is given by:

\begin{equation}
\label{gradient}
\nabla_{\vec{x}}\Big[ h(x_0,x_1,...,x_{n-2},x_{n-1}) \Big] 
			= \Big[ \frac{\partial h}{\partial x_0} , \frac{\partial h}{\partial x_1} , ...\
					\frac{\partial h}{\partial x_{n-2}} , \frac{\partial h}{\partial x_{n-1}} \Big]^T
\end{equation}

\paragraph*{}To build this vector, we need to use the chain rule of multivariate calculus to work \textit{backwards} through the layers of the network to differentiate by layers, hence why the procedure is called \textit{backwards} propagation. To ease mathematical notation, we will redefine the forward pass equation (\ref{mat-vec + act}) into two separate procedures:

\begin{equation}
\label{linear}
\vec{z}^{(l)} = \hat{W}^{(l)} \vec{x}^{(l)} + \vec{b}^{(l)}
\end{equation}
\begin{equation}
\label{activation}
\vec{x}^{(l+1)} = f^{(l)} \big( \vec{z}^{(l)} \big)
\end{equation}

\paragraph*{}We start by differentiating the cost function $J$ with respect to last layer of weights and biases, $\hat{W}^{(L-2})$ and $\vec{b}^{(L-2)}$. The cost, $J$, is a function of the output activations, $\vec{x}^{(L-1)}$ which is in turn a function of $\vec{z}^{L-2}$ in eqn. (\ref{activation}) which is in turn a function of both $\hat{W}^{(L-2})$ and $\vec{b}^{(L-2)}$ in eqn. (\ref{linear}). To account for this layered composition, we multiply each partial derivatives in succession to produce the gradient components that correspond to the needed change in those parameters.

\begin{equation}
\nabla_{W^{(L-2)}} \Big[ J \Big] = 
\frac{\partial J}{\partial x^{(L-1)}} \frac{\partial f^{(L-2)}}{\partial z^{(L-2)}} \frac{\partial z^{(L-2)}}{\partial W^{(L-2)}}
\end{equation}
\begin{equation}
\nabla_{b^{(L-2)}} \Big[ J \Big] = 
\frac{\partial J}{\partial x^{(L-1)}} \frac{\partial f^{(L-2)}}{\partial z^{(L-2)}} \frac{\partial z^{(L-2)}}{\partial b^{(L-2)}}
\end{equation}

Additionally, we compute the gradient with respect to the $L-2$ layer of activations, which will allows for us to move down the to previous sequential layer $L-3$.

\begin{equation}
\nabla_{x^{(L-2)}} \Big[ J \Big] = 
\frac{\partial J}{\partial x^{(L-1)}} \frac{\partial f^{(L-2)}}{\partial z^{(L-2)}} \frac{\partial z^{(L-2)}}{\partial x^{(L-2)}}
\end{equation}

These three equations allow for the model to work backwards through the layers and compute the gradient of the cost function with respect to every paramater in the network. The above equations can be generalized for any layer $l$:

\begin{equation}
\label{dW}
\nabla_{W^{(l)}} \Big[ J \Big] = 
\frac{\partial J}{\partial x^{(l+1)}} \frac{\partial f^{(l)}}{\partial z^{(l)}} \frac{\partial z^{(l)}}{\partial W^{(l)}}
\end{equation}
\begin{equation}
\label{db}
\nabla_{b^{(l)}} \Big[ J \Big] = 
\frac{\partial J}{\partial x^{(l+1)}} \frac{\partial f^{(l)}}{\partial z^{(l)}} \frac{\partial z^{(l)}}{\partial b^{(l)}}
\end{equation}
\begin{equation}
\label{dx}
\nabla_{x^{(l)}} \Big[ J \Big] = 
\frac{\partial J}{\partial x^{(l+1)}} \frac{\partial f^{(l)}}{\partial z^{(l)}} \frac{\partial z^{(l)}}{\partial x^{(l)}}
\end{equation}

\paragraph*{}It is important to know that due to numerical condiderations, this operation is very computationally expensive, and fine details for improving efficiency not considered in the equations above. Additionally, the results from eqns. (\ref{dW}), (\ref{db}), and (\ref{dx}).

@@

\section*{Summary}
